OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-03-16 paper/paper.Rmd
Calculate stats for paper in separate script

```{r deps, message = FALSE}
library(here)
library(mothuR)
library(tidyverse)
dat <- read_tsv(here('results', 'summarized.tsv'))
rel_diff <- function(final, init, percent = TRUE) {
  mult <- if (isTRUE(percent)) 100 else 1
  return((final - init) / init * mult)
}
```

@@

```{r ref_db}
open_fit_mcc <- dat %>% filter(method == 'open', strategy == 'database', tool == 'mothur') %>% pull(mcc_median) %>% median()
opticlust_mcc <- dat %>% filter(method == 'de_novo', tool == 'mothur') %>% pull(mcc_median) %>% median()
mcc_open_fit_db_vs_clust <- rel_diff(opticlust_mcc, open_fit_mcc)

open_fit_gg_mcc <- dat %>% filter(method == 'open', strategy == 'database', tool == 'mothur', ref == 'gg') %>% pull(mcc_median) %>% median()
open_vsearch_mcc <- dat %>% filter(method == 'open', strategy == 'database', tool == 'vsearch', ref == 'gg') %>% pull(mcc_median) %>% median()
mcc_open_fit_db_vs_vsearch <- rel_diff(open_fit_gg_mcc, open_vsearch_mcc)

open_vsearch_sec <- dat %>% filter(method == 'open', strategy == 'database', tool == 'vsearch', ref == 'gg') %>% pull(sec_median) %>% median()
open_fit_sec <- dat %>% filter(method == 'open', strategy == 'database', tool == 'mothur') %>% pull(sec_median) %>% median()
opticlust_sec <- dat %>% filter(method == 'de_novo', tool == 'mothur') %>% pull(sec_median) %>% median()
sec_vsearch_vs_open_fit_db <- rel_diff(open_vsearch_sec, open_fit_sec)
sec_opticlust_vs_open_fit_db <- rel_diff(open_fit_sec, opticlust_sec)
```

```{r de_novo}
dn_vsearch_mcc <- dat %>% filter(strategy == 'de_novo', tool == 'vsearch') %>% pull(mcc_median) %>% median()
dn_vsearch_sec <- dat %>% filter(strategy == 'de_novo', tool == 'vsearch') %>% pull(sec_median) %>% median()
mcc_opticlust_vs_vsearch <- rel_diff(opticlust_mcc, dn_vsearch_mcc)
sec_opticlust_vs_vsearch <- abs(rel_diff(dn_vsearch_sec, opticlust_sec))
```

We assigned blah ref dbs to 3% otus with opticlust. _de novo_ clustering the Greengenes
database, SILVA non-redundant database, or the Ribosomal Database Project (RDP)
[@desantis_greengenes_2006; @quast_silva_2013; @cole_ribosomal_2014].
All clustering was performed at a distance threshold of 0.03 and OTU quality was
evaluated using the MCC as described previously [@westcott_opticlust_2017].
To evaluate reference-based clustering with independent databases, we fit each
dataset to reference OTUs generated by
repeated 100 times.

TODO: use word "map" for what vsearch does, "fit" for what optifit does.

@@

In closed reference mode, OptiFit produced lower quality OTUs than OptiClust by
**X%** when fitting sequences to Greengenes and SILVA, and **Y%** worse when
fitting to RDP.
Only up to **Z%** of query sequences were fit to reference OTUs in closed
reference mode across any dataset/database combination.
OptiFit was able to fit **X** more query sequences to reference OTUs created
with the Greengenes and SILVA databases than with RDP.
VSEARCH was able to fit **W%** more query sequences to the Greengenes reference
than OptiFit in closed reference mode.
In terms of run time, closed reference OptiFit outperformed
OptiClust by **A%** and VSEARCH by **B%**.

>>>>

```{r deps, message = FALSE}
library(here)
library(mothuR)
library(tidyverse)
load(here('results', 'stats.RData'))
```

@@

**TODO: use word "map" for what vsearch does, "fit" for what optifit does.**

**TODO: ref clustering method paragraph here.**
We assigned blah ref dbs to 3% otus with opticlust. _de novo_ clustering the Greengenes
database, SILVA non-redundant database, or the Ribosomal Database Project (RDP)
[@desantis_greengenes_2006; @quast_silva_2013; @cole_ribosomal_2014].
All clustering was performed at a distance threshold of 0.03 and OTU quality was
evaluated using the MCC as described previously [@westcott_opticlust_2017].
To evaluate reference-based clustering with independent databases, we fit each
dataset to reference OTUs generated by
repeated 100 times.

@@

In closed reference mode against databases, OptiFit produced lower quality OTUs
than OptiClust by `r mcc_closed_fit_db_vs_clust` on average.
Fitting sequences to Greengenes and SILVA in closed reference mode performed
similarly with median MCC scores of `r closed_fit_gg_mcc` and 
`r closed_fit_silva_mcc` respectively, while when fitting to RDP the median MCC 
dropped to `r closed_fit_rdp_mcc`.
An average of only `r frac_fit_db` of query sequences were fit to reference
database OTUs in closed reference mode across all dataset/database combinations.
VSEARCH was able to map `r frac_vsearch_vs_fit` more query sequences than
OptiFit to the Greengenes reference database.
In terms of run time, closed reference OptiFit outperformed OptiClust by 
`r sec_closed_fit_db_vs_clust` and VSEARCH by `r sec_closed_fit_db_vs_vsearch`.
